The core transcriptome of mammalian placentas and the divergence of expression with placental shape

### 2016-11-30 manuscript/manuscript.Rnw
fix proteoglycan typo
discuss interdigitation using ancestral state

The core placenta transcriptome is significantly enriched in genes
which are involved in multiple biological pathways, including multiple
pathways related to translation (ribosome %
%
(\Sexpr{placpath("^RIBOSOME","Kegg")})
, translation %
(\Sexpr{placpath("^TRANSLATION","Reactome")})
%
, signal
recognition particle targeting to the rough endoplasmic reticulum %
(\Sexpr{placpath("^SRP DEPENDENT","Reactome")})
%
,
etc.) (Supplemental \cref{tab:core_placenta_pathway_analysis}). Beyond
translation, the next most abundant set of significantly enriched
molecular pathways are related to the migration of the trophoblast
into the endometrium, including the EGFR1 pathway %
(\Sexpr{placpath("^EGFR1","Netpath")})
%
and the Actin Y pathway %
(\Sexpr{placpath("^ACTINY","Biocarta")})
%
. Two other significant pathways, the mPR/Progesterone pathway
and Rho cell motility pathways, feed into the Actin Y pathway as well.
Supplemental \cref{fig:egf_kegg_pathway} shows the EGF pathway in
humans with the median expression in FPKM of genes in the pathway
shown. Notably, many of the genes on the GRB2→MYC pathway are strongly
expressed; their activation leads to an increase in migration and
invasion in response to EGF binding to
EGFR1~\cite{Gupta.ea2016:CellSignalingPathwaysInvolved}. The Actin
cytoskeleton pathway is depicted in Supplemental
\cref{fig:kegg_actin}, which shows that \ac{ITGB1} and its binding
protein \ac{ITGB1BP1} as well as \ac{FN1} are highly expressed, which
activates Rac. This potentially leads to the formation of focal adhesions
and lamellipodia formation through inhibition of GSN and
HSPC300. Other components of cell-cell adhesion, including \ac{F11R}
which forms tight junctions in epithelia, the proteglycans \ac{SDC4} and
\ac{LUM}, and \ac{LAMC1}.

@@

For the degree of intimacy and interdigitation, we saw multiple genes
which were associated with different levels of interdigitiation and
intimacy (Supplemental
\cref{tab:intimacy_genes,tab:interdigitation_genes}) using an
unordered factor approach, but when we used an ordered factor
analysis, none of these genes were retained (Supplemental
\cref{tab:intimacy_genes_polr,tab:interdigitation_genes_polr}). We
suspect this is due to the small number of intermediate
interdigitation and intimacy placentas that are available. Future
analyses of additional placentas with intermediate interdigitation and
intimacy should better illuminate the pathways which are responsible
for these placental morphologies.

>>>>

The core placenta transcriptome is significantly enriched in genes
which are involved in multiple biological pathways, including multiple
pathways related to translation (ribosome %
%
(\Sexpr{placpath("^RIBOSOME","Kegg")})
, translation %
(\Sexpr{placpath("^TRANSLATION","Reactome")})
%
, signal
recognition particle targeting to the rough endoplasmic reticulum %
(\Sexpr{placpath("^SRP DEPENDENT","Reactome")})
%
,
etc.) (Supplemental \cref{tab:core_placenta_pathway_analysis}). Beyond
translation, the next most abundant set of significantly enriched
molecular pathways are related to the migration of the trophoblast
into the endometrium, including the EGFR1 pathway %
(\Sexpr{placpath("^EGFR1","Netpath")})
%
and the Actin Y pathway %
(\Sexpr{placpath("^ACTINY","Biocarta")})
%
. Two other significant pathways, the mPR/Progesterone pathway
and Rho cell motility pathways, feed into the Actin Y pathway as well.
Supplemental \cref{fig:egf_kegg_pathway} shows the EGF pathway in
humans with the median expression in FPKM of genes in the pathway
shown. Notably, many of the genes on the GRB2→MYC pathway are strongly
expressed; their activation leads to an increase in migration and
invasion in response to EGF binding to
EGFR1~\cite{Gupta.ea2016:CellSignalingPathwaysInvolved}. The Actin
cytoskeleton pathway is depicted in Supplemental
\cref{fig:kegg_actin}, which shows that \ac{ITGB1} and its binding
protein \ac{ITGB1BP1} as well as \ac{FN1} are highly expressed, which
activates Rac. This potentially leads to the formation of focal adhesions
and lamellipodia formation through inhibition of GSN and
HSPC300. Other components of cell-cell adhesion, including \ac{F11R}
which forms tight junctions in epithelia, the proteoglycans \ac{SDC4} and
\ac{LUM}, and \ac{LAMC1}.

@@

For the degree of intimacy and interdigitation, we saw multiple genes
which were associated with different levels of interdigitiation and
intimacy (Supplemental
\cref{tab:intimacy_genes,tab:interdigitation_genes}) using an
unordered factor approach, but when we used an ordered factor
analysis, none of these genes were retained (Supplemental
\cref{tab:intimacy_genes_polr,tab:interdigitation_genes_polr}). We
suspect that this is due to the fact that the ancestral state of all
eutherians is a labyrinthine, hemochorial
placenta~\cite{Wildman.ea2006:Evolutionofmammalianplacenta}. This
implies that the transcriptome of intermediately invasive or
interdigitated placentas need not be intermediate between the
transcriptomes of less intimate (folded/lamellar) or interdigitated
placentas (epitheliochorial) and most intimate (hemochorial) or
interdigitated placentas (labyrinthine) because these morphologies
have evolved independently. An alternative is that the limited number
of species with placentas of intermediate interdigitation and intimacy
in this study limits the ability to detect genes which whose
expression correlates with degree of intimacy and interdigitation.
Future analyses of additional placentas with intermediate
interdigitation and intimacy should better illuminate the pathways
which are responsible for these placental morphologies.
